\begin{abstract}
We propose two parameter-efficient architectural additions to hybrid state-space/attention language models --- \emph{hormone routing}, a token-conditional router that injects a gated combination of frozen direction vectors into the residual stream after each attention block, and \emph{intracellular attention}, a slot-attention mechanism over the $\dstate$ memory cells of a selective state-space block --- and use them to investigate the relationship between architectural inductive bias and the optimisation objective. Both mechanisms use a zero-initialised output gate so that the unmodified hybrid baseline is recovered exactly at initialisation. We pretrain a 125M-parameter Mamba/MQA hybrid on FineWeb-Edu under a fixed 2B-token budget and report a striking decoupling between the routing module's representational organisation and its effect on the loss objective. The hormone-routing module develops semantically-coherent per-genre routing specialisation across every variant we examine --- including controls that replace the extracted directions with random unit vectors, with vectors extracted from a randomly-initialised model, or that freeze the output gate at $1.0$. Maximum pairwise $L_1$ distance between genre routing distributions ranges from $0.6$ to $1.6$ (uniform routing $= 0.0$, disjoint $= 2.0$), with the variant of \emph{highest} specialisation strength being the one whose gate the model cannot modulate. Yet \textbf{specialisation strength does not correlate with loss}: the four properly-initialised variants converge to identical validation perplexity ($26.13$). To probe whether this loss-invariance reflects a fundamental ceiling on the mechanism's contribution or a regime the model actively absorbs, we train a sixth variant with a forced non-zero gate initialisation and a non-trainable $3\times$ amplification on the injection. The model responds by drifting the gates downward and shrinking the per-hormone magnitudes, but the residual perturbation it cannot suppress --- a roughly $2\times$ amplification of the unforced regime --- does not damage validation loss at the same training point. The picture that emerges is of a \emph{homeostatic envelope}: within it, the gated injection is absorbed without affecting predictions, and the routing module's representational organisation develops freely. The envelope is wider than the unforced variants naturally select, and the model expresses a preference for smaller perturbations within it without requiring them. The envelope can still be broken --- by introducing the routing during base pretraining rather than as a continuation from a stabilised checkpoint --- which damages the loss by $+2.07$ PPL even though the same routing module develops similar specialisation in this regime as in any other. The loss-relevant axis is the training schedule, not the perturbation magnitude. Both contributions add fewer than $1\%$ to the base model's parameter count and are released as open-source code.
\end{abstract}
