\section{Introduction}
\label{sec:intro}

Hybrid architectures that interleave selective state-space mixers~\citep{gu2023mamba} with attention~\citep{vaswani2017attention} have emerged as a strong alternative to pure-attention transformers, combining the linear-time long-range modelling of structured state-space models (SSMs) with the precise content-based retrieval that attention provides~\citep{lieber2024jamba,glorioso2024zamba,park2024empirical}. These hybrids retain attention's expressivity at the locations where it matters most, while relegating the bulk of the sequence mixing to the more parameter-efficient SSM path.

We propose two parameter-efficient architectural additions to this family and use them to investigate the relationship between architectural inductive bias and the optimisation objective. The two additions target distinct loci. \emph{Intracellular attention} concerns the SSM's \emph{internal} structure: the selective state $h_t \in \mathbb{R}^{\dinner \times \dstate}$ of a Mamba block is best read as $\dinner$ parallel channels, each carrying a $\dstate$-dimensional memory; within each channel the slots evolve independently and only ever interact through the linear readout $y_t = h_t C_t^\top$. We replace this linear contraction with full content-dependent attention over the slots. \emph{Hormone routing} concerns the residual stream \emph{between} blocks: it injects a token-conditional combination of a small bank of frozen direction vectors --- extracted from the partially-trained base model itself via document-style contrast pairs~\citep{turner2023activation,rimsky2024steering,zou2023representation} --- after each attention block, gated by a learned scalar.

Both mechanisms share a design discipline. Each adds fewer than $1\%$ to the base parameter count. Each is gated such that, at initialisation, the modification is an exact no-op and the hybrid backbone is recovered bit-for-bit. Each is structurally orthogonal to the other.

Our empirical contribution is qualitatively different from the contribution we initially set out to make. Across every hormone-routing variant we tested at the $125$M-parameter, $2$B-token scale --- with extracted directions, with random unit vectors, with the output gate frozen at $1.0$, and with directions extracted from a randomly-initialised model --- the router converges to a strongly genre-discriminative routing distribution (max pairwise $L_1$ distance between routing distributions across six text genres: $0.6$--$1.6$, vs $0.0$ for uniform routing). The specialisation is qualitatively interpretable in the variants where the directions carry semantic labels (in the extracted-direction variant, news routes preferentially to the urgency hormone, fiction to curiosity, code to skepticism, urgent text to warmth) and is structurally analogous in the variants where they do not. Yet none of these variants moves the validation perplexity needle: all properly-initialised hormone variants achieve identical validation PPL ($26.13$), matching the unmodified base. Only the variant that introduces hormone routing during base pretraining --- rather than as a continuation from a stabilised checkpoint --- changes the loss, and it does so by degrading it ($28.20$ PPL).

We interpret this finding as an instance of a broader phenomenon: \textbf{the model's architectural channels can develop structured representational organisation that is not visible in the loss}. The router learns the genre-conditional distribution it would need in order to use the hormone bank meaningfully, but the bounded perturbation magnitude --- gate $g$ converges to $0.23$ at the upper attention layer, $\approx 0$ at the lower one --- is too small to alter the model's predictions at this scale. Whether this represents a fundamental ceiling on the mechanism's contribution, or merely a capacity-bottlenecked regime in which the residual stream is too low-dimensional to carry useful gradient signal, is a question we partially address with a forced-engagement variant (Section~\ref{sec:experiments}).

Our contributions are as follows.

\begin{itemize}
    \item \textbf{Empirical: a clean decoupling between routing specialisation and loss.} Across five hormone-routing variants pretrained on a $125$M-parameter Mamba/MQA hybrid for $2$B tokens, the routing module develops strongly genre-discriminative routing distributions in every condition we test --- including controls that replace the extracted directions with random unit vectors and that freeze the gate at $1.0$ --- with maximum pairwise $L_1$ distance between genre routing distributions in the range $0.6$--$1.6$ (uniform $= 0.0$, disjoint $= 2.0$). Yet the variant with the \emph{highest} specialisation strength achieves the same validation perplexity as the variant with the \emph{lowest} and as the unmodified baseline. \textbf{Specialisation strength does not predict downstream loss}, and the model is invariant to the routing module's content (extracted vs random vs fixed-gate) but not to its training-schedule timing (continuation vs from-scratch).

    \item \textbf{Architectural: hormone routing.} The mechanism that produces the above finding is a parameter-efficient routing channel: a per-token softmax-gated linear combination of a small bank of frozen direction vectors, added to the residual stream after each attention block, multiplied by a learned scalar gate. The router is small (under $0.01\%$ of the base model's parameters) and the bank is non-trainable. We further propose a \emph{document-style contrast-pair extraction protocol} that recovers behavioural directions from the base model itself --- without instruction-tuning data or annotated preferences --- and discuss the methodological reasons for preferring it to instruction-style pairs in the pretraining setting.

    \item \textbf{Architectural: intracellular attention.} A second proposed mechanism, structurally orthogonal to hormone routing, places content-dependent attention inside the SSM scan to let the $\dstate$ state slots attend to each other before the linear readout. The mechanism preserves the SSM's recurrent form, adds parameters proportional only to a small bottleneck dimension $\dattn \ll \dinner$, and is gated identically (zero-init, learnable) so that the unmodified scan is recovered at initialisation.

    \item \textbf{Forced-engagement test.} A sixth variant initialises the gate at a non-zero value and amplifies the injection by a constant multiplier, producing an effective initial perturbation several times larger than the largest converged gate of the unforced variants. This isolates whether the loss-decoupling reflects a fundamental property of the mechanism or a capacity-bottlenecked regime in which the model can absorb only small perturbations.
\end{itemize}

The rest of the paper is structured as follows. Section~\ref{sec:related} situates the work against the SSM, hybrid, representation-engineering, and routing literatures. Section~\ref{sec:arch} gives the hybrid backbone. Sections~\ref{sec:intracellular} and~\ref{sec:hormone} present the two architectural contributions in turn. Section~\ref{sec:experiments} reports the empirical study. Section~\ref{sec:analysis} characterises what the router learns and discusses the implications for representation-level organisation independent of objective.
