\section{Architecture}
\label{sec:arch}

We use a hybrid Mamba/MQA backbone of $L$ residual layers with a fixed 5:1 SSM-to-attention layer pattern (i.e.\ five SSM layers between every attention layer, with a final attention layer if $L \bmod 6 \neq 0$). All layers share the same hidden width $\dmodel$. Layer norms are RMSNorm~\citep{zhang2019root}; word embeddings and the language-model head share weights.

\paragraph{SSM layers.} Each SSM layer applies pre-norm and a Mamba-1 selective SSM block~\citep{gu2023mamba}, with no separate channel-mixing MLP. Concretely, given the layer input $x \in \mathbb{R}^{B \times T \times \dmodel}$, we project to twice the inner width, $[x', z'] = W_{\text{in}} \, \text{RMSNorm}(x)$ with $x', z' \in \mathbb{R}^{B \times T \times \dinner}$ and $\dinner = 2\dmodel$, apply a causal depthwise convolution to $x'$, and run the input through a selective scan parameterised by an input-dependent discretisation step $\Delta_t$, input matrix $B_t \in \mathbb{R}^{\dstate}$, and readout matrix $C_t \in \mathbb{R}^{\dstate}$. The state $A \in \mathbb{R}^{\dinner \times \dstate}$ is parameterised in log-space as a per-channel diagonal. The scan output is gated multiplicatively by SiLU$(z')$ and projected back to width $\dmodel$:
\begin{align}
h_t &= \Delta_t \, A \odot h_{t-1} + \Delta_t \, B_t \odot x_t, \\
y_t &= h_t \, C_t^\top + D \odot x_t,
\end{align}
with $h_t \in \mathbb{R}^{\dinner \times \dstate}$ and $D \in \mathbb{R}^{\dinner}$ a per-channel skip. Section~\ref{sec:intracellular} replaces the linear contraction $h_t C_t^\top$ with attention over the $\dstate$ slots.

\paragraph{Attention layers.} Each attention layer applies pre-norm followed by multi-query attention~\citep{shazeer2019fast} with a single shared key/value head, rotary positional embeddings on the query and key projections, and a SwiGLU~\citep{shazeer2020glu} channel-mixing MLP of inner width $d_{\text{ff}} = \lceil \nicefrac{8}{3}\dmodel \rceil$ rounded to a multiple of $64$. The hormone routing layer of Section~\ref{sec:hormone} adds a third residual addition after the MLP residual:
\begin{align}
x &\leftarrow x + \text{Attn}(\text{RMSNorm}(x)),\\
x &\leftarrow x + \text{MLP}(\text{RMSNorm}(x)),\\
x &\leftarrow x + \text{Hormone}(x).
\end{align}

\paragraph{Configuration ranges.} We work primarily with a 350M-parameter configuration ($\dmodel = 1536$, $L = 18$, $\dinner = 3072$, $\dstate = 16$, $\dattn = 16$, twelve query heads, one shared KV head, head dim $128$); a 125M tiny variant for fast ablation; and a 770M medium variant. Layer counts and widths for each configuration are summarised in Table~\ref{tab:configs}.

\begin{table}[t]
\centering
\caption{Backbone configurations. All sizes share $\dstate = 16$, $\dattn = 16$, head dim $128$, MQA, 5:1 SSM:Attn pattern.}
\label{tab:configs}
\begin{tabular}{lccccc}
\toprule
Name & Params & $\dmodel$ & $L$ & $n_{\text{heads}}$ & SSM/Attn layers \\
\midrule
\textsc{tiny}   & $\sim$125M & 1024 & 12 & 8  & 10/2 \\
\textsc{small}  & $\sim$350M & 1536 & 18 & 12 & 15/3 \\
\textsc{medium} & $\sim$770M & 2048 & 24 & 16 & 20/4 \\
\bottomrule
\end{tabular}
\end{table}
