\section{Analysis}
\label{sec:analysis}

\subsection{What does the hormone router learn?}

We probe the trained \textsc{HR} model by recording, for each token in a held-out validation batch, the router weights $w(h) \in \Delta^{\nhormones - 1}$ at every attention layer. Figure~\TBD{X} visualises the average per-token routing distribution over a corpus of varied genres (news, fiction, code, instructional text, casual dialogue). \TBD{Narrative: do the hormones specialise per-genre, per-token-type, per-position? Provide the qualitative pattern.}

\paragraph{Magnitude evolution.} The per-hormone learned magnitudes $m_i$ were initialised at $1.0$ across all hormones. Figure~\TBD{Y} plots their value trajectories across Stage~2 training. \TBD{Narrative: which hormones grew, which shrank, how monotone are the trajectories.}

\paragraph{Gate trajectories.} Each attention layer's gate $g$ was initialised at $0$. Figure~\TBD{Z} shows the gate values per layer over Stage~2. \TBD{Narrative: do all layers eventually open the gate? Are upper layers earlier? Late layers later?}

\subsection{What does intracellular attention learn?}

For the trained \textsc{IA} model, we record the slot-to-slot attention pattern $\alpha_t \in \mathbb{R}^{\dstate \times \dstate}$ at each scan step on which intracellular attention fires, averaged over the validation set. We plot the average attention pattern at three depths (early, middle, late SSM layer). \TBD{Narrative: do early layers diffuse uniformly across slots, while late layers exhibit a sparse pattern? Is the gate $g$ open at all layers or only some?}

\paragraph{Slot-pair specialisation.} If intracellular attention is doing useful work, we expect specific slot pairs to attend to one another consistently across inputs. Figure~\TBD{W} reports the entropy of $\alpha_t$ averaged across positions, layers, and inputs. \TBD{Narrative: lower entropy than uniform indicates the mechanism has chosen specific slot interactions.}

\subsection{Limitations}

\paragraph{Scale.} Our experiments are at the 125M-parameter scale on a 2B-token budget. While these are sufficient to observe statistically meaningful differences between variants, scaling laws~\citep{hoffmann2022training} predict that the relative impact of architectural modifications can change with model size. We cannot rule out that the contributions reported here either grow or shrink in importance at larger scale.

\paragraph{Pretraining-only.} We do not perform supervised fine-tuning or RLHF. The hormone routing mechanism is in principle compatible with these stages --- the routing weights and magnitudes can be further updated, and re-extraction can target newly-relevant behavioural axes --- but we leave that investigation to future work.

\paragraph{Single-architecture study.} Both contributions are designed for the Mamba/MQA hybrid family. Intracellular attention generalises naturally to any SSM with a state shape that admits the slot interpretation (S4, S5, Mamba-2). Hormone routing is architecture-agnostic and could be applied to a pure-transformer base, but we have not validated this empirically.
