\section{Conclusion}
\label{sec:conclusion}

We have presented two parameter-efficient architectural additions to hybrid state-space and attention language models. \emph{Hormone routing} is a token-conditional router that injects a gated combination of frozen direction vectors into the residual stream after each attention block. \emph{Intracellular attention} is a slot-attention mechanism placed inside the SSM scan that lets the otherwise-independent state slots interact via content-dependent attention before the linear readout. Both add fewer than $1\%$ of trainable parameters and are gated such that the unmodified hybrid backbone is recovered exactly at initialisation.

Our central empirical finding is a clean decoupling between two quantities that interpretability work typically assumes are correlated. The hormone-routing module develops semantically-coherent per-genre routing specialisation across every variant we examined --- including controls that replace the extracted directions with random unit vectors and that freeze the output gate at $1.0$ --- with maximum pairwise $L_1$ distance between genre routing distributions in the range $0.6$--$1.6$ (uniform $= 0$, disjoint $= 2$). The variant with the \emph{highest} specialisation strength is one that the model cannot modulate (\textsc{HR-fixedgate}, gate frozen at $1.0$). Yet at $125$M parameters and $2$B tokens, none of this organisation produces a measurable contribution to the loss objective: all properly-initialised variants converge to identical validation perplexity. \textbf{Specialisation strength does not correlate with loss.} The forced-engagement variant (\textsc{HR-forced}) constrains the interpretation of this decoupling. By initialising the gate at $0.2$ and applying a non-trainable $\alpha = 3.0$ amplifier, we force the model into an effective-injection regime $\approx 2\times$ that of any unforced variant. The model responds by drifting the gates downward during training and shrinking the per-hormone magnitudes --- expressing a preference for smaller perturbations through every parameter it can modulate --- yet the residual it cannot suppress (because $\alpha$ is non-trainable) does \emph{not} damage validation loss at the same training point as the unforced variants. The architecture-and-data system has a \emph{homeostatic envelope} for additive routing perturbations: within it, the gated injection is absorbed without affecting predictions, and the routing module's representational organisation develops freely. The envelope is wider than the unforced variants naturally select, and the loss-relevant axis on which the mechanism can break the model is the training schedule (\textsc{HR-noext}: introducing the perturbation before the homeostatic capacity is in place), not the perturbation magnitude itself within the envelope.

The mechanism is bounded --- random vectors and frozen gates yield base-level loss --- but sensitive to the training schedule: introducing hormone routing during base pretraining (rather than as a continuation from a stabilised checkpoint) actively degrades performance, even though the router still develops a discriminative routing distribution.

\paragraph{Future directions.} A fused CUDA kernel for intracellular attention would close the gap to the standard Mamba scan and remove the practical barrier to applying the mechanism at scale; we have established the architecture and a preliminary engagement signal, but the full empirical case requires the kernel. The hormone-extraction protocol, currently performed once at the $1$B-token milestone, could be made continuous --- updating $V$ on a slow schedule throughout training --- which would let the bank track the model's evolving representations. Most importantly, the routing-organisation-without-loss-relevance finding should be tested at larger scale: whether the bounded-engagement regime we observe at $125$M parameters reflects a fundamental property of the mechanism or a capacity-bottlenecked phenomenon depends on how the residual stream's dimensionality interacts with the perturbation magnitude. The forced-engagement variant we report begins this investigation but does not settle it.

A broader research programme suggested by this work is the systematic study of \emph{introspective architectural channels}: components that develop structured internal organisation along axes the architecture itself defines, observable to the model and to interpretation, but decoupled --- at least at small scale --- from the optimisation objective. The hormone-routing module is one example. Whether such channels become loss-relevant at scale, and what conditions cause them to do so, remains an open question we have only begun to examine.
